---
title: "Assignment 3: Data Exploration"
author: "Hiranmayi Nataraj"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `HiranmayiNataraj_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
install.packages('here')
install.packages("lubridate")
install.packages("tidyverse")
```
```{r}
library(here)
litter <- read.csv(
file = here('Data','Raw','NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
stringsAsFactors = T
)

neonids <- read.csv(
file = here('Data','Raw','ECOTOX_Neonicotinoids_Insects_raw.csv'),
stringsAsFactors = T
)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: The interest in the ecotoxicology of neonicotinoids on insects is driven by their widespread use in agriculture and their potential impact on both target and non-target insect species. Neonicotinoids are highly effective insecticides,these chemicals can persist in the environment, potentially causing long-term exposure and bioaccumulation in ecosystems. Understanding their impact is crucial for assessing environmental risks, informing regulatory decisions, and developing sustainable pest management strategies.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:This material plays a vital role in nutrient cycling, providing essential elements for plant growth as it decomposes. It also serves as important habitat for various organisms, from microbes to small mammals, enhancing biodiversity. Woody debris contributes significantly to carbon sequestration, helping mitigate climate change. Additionally, it aids in soil stability, moisture retention, and forest regeneration. Research at sites like the Niwot Ridge LTER station provides valuable insights into these processes, informing forest management strategies and our understanding of forests' role

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Litter is collected in elevated 0.5 m² PVC traps for material with a diameter <2 cm and length <50 cm. Fine woody debris, with a diameter <2 cm and length >50 cm, is collected in larger ground traps (3m x 0.5m rectangular areas).
 2.Litterfall and fine woody debris are sampled in randomly selected tower plots, with each plot receiving one to four trap pairs, depending on the size of the plot. Trap placement can either be random or targeted based on the type and cover of vegetation at the site.
 3. Ground traps are sampled once annually, while elevated traps are sampled more frequently, with a target of every 1-2 weeks in deciduous forests during leaf senescence, and once every 1-2 months in evergreen sites. Sampling may be paused for up to six months in deciduous sites during the dormant season.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
summary(neonids)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
summary(neonids$Effect)
sorted_effects <- sort(table(neonids$Effect), decreasing = TRUE)
sorted_effects
```

> Answer: The most common effects, such as population, mortality, behavior, and reproduction, are of significant interest because they directly relate to the survival and health of species within ecosystems. Population dynamics and mortality rates are crucial for understanding the long-term viability of species, especially in response to environmental stressors like toxins or habitat disruption. Changes in behavior, including feeding behavior, often serve as early indicators of environmental stress, while disruptions in reproduction can signal broader ecological issues and affect species sustainability.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
summary(neonids$Species.Common.Name)
sorted_species <- sort(table(neonids$Species.Common.Name), decreasing = TRUE)
head(sorted_species, 6)
```

> Answer: These species—honey bees, bumble bees, and parasitic wasps—are all crucial pollinators or play significant roles in pest control, making them integral to both ecosystem health and agricultural productivity. Their sensitivity to neonicotinoid pesticides, combined with their ecological and economic importance, makes them key species for studying the broader impacts of these chemicals.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
class(neonids$Conc.1..Author.)
df <- table(neonids$Conc.1..Author.)
df
```

> Answer:The class is factor. The Conc.1..Author. column is a factor because it contains concentrations with non-numeric symbols like <, >, and ~, which R treats as categorical values.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
library(ggplot2)

pub_year <- (neonids$Publication.Year)

ggplot(neonids, aes(x = pub_year)) +
  geom_freqpoly(binwidth = 1, color = "blue")
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
neonids$Test.Location <- factor(neonids$Test.Location)
test_l <- neonids$Test.Location
ggplot(neonids, aes(x = pub_year, color = test_l)) +
  geom_freqpoly(binwidth = 1)

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:Field natural, field artificial, lab. Yes they do differ over time

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
library(ggplot2)
library(dplyr)

endpoint_counts <- count(neonids, Endpoint)
endpoint_counts <- arrange(endpoint_counts, desc(endpoint_counts$n))

ggplot(endpoint_counts, aes(x = reorder(Endpoint, -n), y = n)) +
  geom_bar(stat = "identity", fill = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: Two most common are NOEL and LOEL.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(litter$collectDate)
litter$collectDate <- as.Date(litter$collectDate, format = "%Y-%m-%d")
class (litter$collectDate)
print(unique(litter$collectDate))
```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique_plots <- unique(litter$plotID)

#count of unique plots
unique_len <- length(unique_plots)
unique_len

summary_plot <- summary(litter$plotID)
summary_plot
```

> Answer: 12. unique() returns only the distinct values in the column , summary() provides a more descriptive statistical summary of the data

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
functional_group_counts <- table(litter$functionalGroup)

functional_group_df <- as.data.frame(functional_group_counts)

library(ggplot2)

ggplot(functional_group_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
library(ggplot2)

ggplot(litter, aes(x = functionalGroup, y = dryMass)) +
  geom_boxplot(fill = "steelblue", color = "black") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```
```{r}
# Violin plot
ggplot(litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin(fill = "steelblue", color = "black") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The boxplot is more effective in this case because it clearly summarizes the central tendency, spread, and outliers of dryMass by functional group, making it easier to compare groups.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles
