---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Hiranmayi Nataraj"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse) 
install.packages("agricolae")
library(agricolae)  
library(lubridate)
library(ggplot2)

getwd()

setwd("~/EDA_Spring2025/Data/Raw")

chem_data <- read_csv("NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

chem_data <- chem_data %>%
  mutate(sampledate = ymd(sampledate))

#2

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature does not change with depth.
Ha: Mean lake temperature changes with depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
july_lake_data <- chem_data %>%
  filter(month(sampledate) == 7) %>%  
  select(lakename, year4, daynum, depth, temperature_C) %>%  
  drop_na()

head(july_lake_data)
unique(july_lake_data$lakename)

#5
ggplot(july_lake_data, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.6, color = "blue") +  # Scatter points with transparency
  geom_smooth(method = "lm", color = "red", se = TRUE) +  # Linear model with confidence interval
  scale_y_continuous(limits = c(0, 35)) +  # Limit temperature from 0 to 35°C
  labs(
    title = "Lake Temperature vs. Depth (July)",
    x = "Depth (m)",
    y = "Temperature (°C)",
  ) +
  mytheme  

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
As depth increases, lake temperature decreases significantly i.e the red line shows a downward trend. But the spread of points suggests that the rate of temperature decrease might be steeper at shallow depths and more gradual at deeper depths. The lake temperature strongly decreases with depth, but the spread of data suggests a potentially non-linear trend.

7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

temp_depth_lm <- lm(temperature_C ~ depth, data = july_lake_data)
summary(temp_depth_lm)

cor(july_lake_data$depth, july_lake_data$temperature_C, use = "complete.obs")
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
* The Multiple R-squared = 0.7387 indicates that 73.87% of the variability in lake temperature is explained by depth.This suggests a strong negative relationship between temperature and depth.
* The p-value (<2.2e-16) is extremely small, indicating that the relationship between temperature and depth is highly statistically significant.
*The Residual Standard Error (3.835) suggests that the model’s predictions typically deviate by about 3.835°C from the actual temperature values.
*Given the strong significance (p < 0.001), we reject the null hypothesis (H₀) that depth does not affect temperature.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

lake_temp_model <- lm(data = july_lake_data, temperature_C ~ year4 + daynum + depth)
summary(lake_temp_model)

library(corrplot)

lake_corr_matrix <- july_lake_data %>%
  select(year4, daynum, depth, temperature_C) %>%
  cor(use = "complete.obs")

# Correlation plot
par(mar = c(7, 7, 7, 7)) 
corrplot(lake_corr_matrix, method = "ellipse", tl.cex = 0.7)

##Choose a model by AIC in a Stepwise Algorithm
step(lake_temp_model)
summary(lake_temp_model)

#10

final_temp_model <- lm(data = july_lake_data, temperature_C ~ year4 + daynum + depth)
summary(final_temp_model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The AIC-selected model includes year4, daynum, and depth. This model improves variance from ~73.87% to 74.12%, meaning adding year4 and daynum explains a little more variability in temperature. It improves upon the depth-only model by ~0.25%, suggesting depth is the key factor, but time-based variables (year4 and daynum) add minor predictive value.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
anova_model <- aov(data = july_lake_data, temperature_C ~ lakename)
summary(anova_model)

lm_model <- lm(data = july_lake_data, temperature_C ~ lakename)
summary(lm_model)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
Anova : F-statistic = 50 and p-value < 2e-16 (very small) 
This confirms significant temperature differences exist between lakes
Linear model : All p-values are significant (p < 0.05) - Each lake listed has a statistically significant difference in temperature compared to the reference.R² = 0.0395 - The linear model explains only 3.95% of the variability in lake temperature.
While there is a significant difference in temperatures between lakes, the lake name alone does not explain much of the total variability (other factors like depth, time of year, or lake-specific conditions likely play a larger role).


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(july_lake_data, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +  
  scale_y_continuous(limits = c(0, 35)) + 
  labs(
    title = "Lake Temperature vs. Depth (July)",
    x = "Depth (m)",
    y = "Temperature (°C)",
    color = "Lake Name"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "right"
  )

ggplot(july_lake_data, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, size = 2) +  # Improve visibility of points
  geom_smooth(method = "lm", se = FALSE) +  # Add separate trendlines
  scale_y_continuous(limits = c(0, 35)) +  # Restrict y-axis range
  labs(
    title = "Lake Temperature vs. Depth (July)",
    x = "Depth (m)",
    y = "Temperature (°C)",
    color = "Lake Name"
  ) +
  theme_minimal() +  # Use clean theme
  theme(
    legend.position = "bottom",  # Move legend for better spacing
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x labels if needed
  ) +
  facet_wrap(~lakename)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
anova_model <- aov(temperature_C ~ lakename, data = july_lake_data)
summary(anova_model)

tukey_results <- TukeyHSD(anova_model)
print(tukey_results)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Peter Lake has the same mean temperature as: East Long Lake, Paul Lake, Tuesday Lake and Ward Lake. No lake has a mean temperature that is significantly different from all other lakes (all p-values are above 0.05).
 
17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we want to compare only Peter Lake and Paul Lake to determine whether their mean temperatures are significantly different, another statistical test we can use is the independent (unpaired) t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
crampton_ward_data <- july_lake_data %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

t_test_result <- t.test(temperature_C ~ lakename, data = crampton_ward_data)
print(t_test_result)
```

>Answer: 
